\documentclass{article}

\usepackage{amsmath, amssymb}

\title{Continued Fraction Evaluation}
\author{}
\date{}

\begin{document}

\maketitle

\section{Fractional linear transformations}

A fractional linear transformation (or M\"obius transformation) is a function of the form
\begin{equation}
  f(z) = \frac{az + b}{cz + d},
\end{equation}
which we represent by the $2 \times 2$ matrix
\begin{equation}
  M = \begin{pmatrix} a & b \\ c & d \end{pmatrix}.
\end{equation}
We write $M \cdot z$ for the action of $M$ on $z$ by the corresponding fractional linear transformation:
\begin{equation}
  \begin{pmatrix} a & b \\ c & d \end{pmatrix} \cdot z = \frac{az + b}{cz + d}.
\end{equation}
The key property is that composition of fractional linear transformations corresponds to matrix multiplication: if $M_1$ and $M_2$ are two such matrices, then
\begin{equation}
  M_1 \cdot (M_2 \cdot z) = (M_1 M_2) \cdot z.
\end{equation}
Because matrix multiplication is associative, this gives us freedom to group and reorder the computation of a composition of fractional linear transformations in any way we like.

This action can be realized concretely by identifying a scalar $z$ with the homogeneous column vector $\binom{z}{1}$ and letting $M$ act by matrix-vector multiplication on the right:
\begin{equation}
  \begin{pmatrix} a & b \\ c & d \end{pmatrix} \begin{pmatrix} z \\ 1 \end{pmatrix} = \begin{pmatrix} az + b \\ cz + d \end{pmatrix},
\end{equation}
which represents the value $(az + b)/(cz + d)$. The distinguished points correspond to
\begin{equation}
  0 \;\longleftrightarrow\; \begin{pmatrix} 0 \\ 1 \end{pmatrix}, \qquad
  \infty \;\longleftrightarrow\; \begin{pmatrix} 1 \\ 0 \end{pmatrix},
\end{equation}
so acting on $\infty$ reads off the first column of the matrix ($a/c$), and acting on $0$ reads off the second column ($b/d$).

\section{Polynomials}

Consider a polynomial
\begin{equation}
  p(x) = a_0 + a_1 x + a_2 x^2 + \cdots + a_n x^n.
\end{equation}
Horner's method rewrites this as a nested evaluation
\begin{equation}
  p(x) = a_0 + x\bigl(a_1 + x\bigl(a_2 + \cdots + x \cdot a_n \cdots\bigr)\bigr).
\end{equation}
Each level of the nesting corresponds to a fractional linear transformation $z \mapsto a_k + xz$, represented by the matrix
\begin{equation}
  L_k = \begin{pmatrix} x & a_k \\ 0 & 1 \end{pmatrix}.
\end{equation}
The value of the polynomial is then given by acting on $\binom{0}{1}$ on the right:
\begin{equation}
  (L_0 L_1 \cdots L_n) \begin{pmatrix} 0 \\ 1 \end{pmatrix} = \begin{pmatrix} p(x) \\ 1 \end{pmatrix}.
\end{equation}
Writing
\begin{equation}
  L_0 L_1 \cdots L_k = \begin{pmatrix} x^{k+1} & p_k(x) \\ 0 & 1 \end{pmatrix},
\end{equation}
where $p_k(x) = a_0 + a_1 x + \cdots + a_k x^k$, we see that the accumulated product directly tracks the partial sums of the polynomial.

Because each $L_k$ has bottom row $(0, 1)$, the denominator of the homogeneous result is always $1$: there is no need to divide at the end. We can extract the polynomial value as a single number by also acting with a row vector on the left:
\begin{equation}
  p(x) = \begin{pmatrix} 1 & 0 \end{pmatrix} L_0 L_1 \cdots L_n \begin{pmatrix} 0 \\ 1 \end{pmatrix}.
\end{equation}
This sandwiched form is purely a product of matrices and vectors, with no homogeneous-to-scalar conversion needed. Associativity now applies to the entire expression: we may group and partition the product in any way we like, and each sub-product is an ordinary matrix or scalar multiplication.

Accumulating the matrix product from right to left gives the \emph{backward recurrence}
\begin{equation}
  p(x) = a_0 + x\bigl(a_1 + x\bigl(a_2 + \cdots + x \cdot a_n\bigr)\cdots\bigr),
\end{equation}
which is the well-known Horner's method (or nested multiplication). Accumulating from left to right gives the \emph{forward recurrence}
\begin{equation}
  p(x) = \bigl(\cdots\bigl((a_n \cdot x + a_{n-1}) \cdot x + a_{n-2}\bigr) \cdot x + \cdots\bigr) \cdot x + a_0,
\end{equation}
which processes the coefficients in the opposite order, starting from the leading term. Because matrix multiplication is associative, both orderings---and any other grouping of the matrix product---yield the same result.

\subsection*{Estrin's method}

Beyond left-to-right and right-to-left accumulation, we can also group the matrix product into pairs and multiply in a balanced binary tree. Pairing adjacent matrices gives
\begin{equation}
  L_k L_{k+1}
  = \begin{pmatrix} x & a_k \\ 0 & 1 \end{pmatrix}
    \begin{pmatrix} x & a_{k+1} \\ 0 & 1 \end{pmatrix}
  = \begin{pmatrix} x^2 & a_k + a_{k+1} x \\ 0 & 1 \end{pmatrix}.
\end{equation}
The nontrivial content of each pair product is just the linear combination $a_k + a_{k+1} x$, plus the power $x^2$ which is shared across all pairs. All pairs can be computed simultaneously. At the next level, each paired product has the form $\bigl(\begin{smallmatrix} x^2 & q_j \\ 0 & 1 \end{smallmatrix}\bigr)$, so multiplying two of them gives $\bigl(\begin{smallmatrix} x^4 & q_j + x^2 q_{j+1} \\ 0 & 1 \end{smallmatrix}\bigr)$: combine pairs using $x^2$ and compute $x^4$ for the next round.

This is exactly Estrin's method, which groups a polynomial into linear pieces that can be evaluated in parallel and then combines them level by level. The correspondence works cleanly because the bottom row stays $(0, 1)$ at every level---upper triangular times upper triangular is upper triangular---so the matrix product never does any work beyond what the scalar Estrin scheme does.

\subsection*{Constant coefficients and geometric series}

When all coefficients are equal, $a_k = a$, every matrix is the same:
\begin{equation}
  L = \begin{pmatrix} x & a \\ 0 & 1 \end{pmatrix},
\end{equation}
and the polynomial $p(x) = a(1 + x + \cdots + x^n)$ is computed by the matrix power $L^{n+1}$. The two standard methods for computing a matrix power---diagonalization and repeated squaring---each correspond to a well-known technique for summing a geometric series.

\paragraph{Diagonalization.}
$L$ has eigenvalues $x$ and $1$, so $L = P \bigl(\begin{smallmatrix} x & 0 \\ 0 & 1 \end{smallmatrix}\bigr) P^{-1}$ with
\begin{equation}
  P = \begin{pmatrix} 1 & a \\ 0 & 1-x \end{pmatrix}, \qquad
  P^{-1} = \frac{1}{1-x}\begin{pmatrix} 1-x & -a \\ 0 & 1 \end{pmatrix}.
\end{equation}
Computing $L^{n+1} = P D^{n+1} P^{-1}$ gives
\begin{equation}
  L^{n+1} = \begin{pmatrix} x^{n+1} & a\,\dfrac{1-x^{n+1}}{1-x} \\[6pt] 0 & 1 \end{pmatrix},
\end{equation}
and reading off the upper-right entry recovers the standard closed form for a finite geometric series:
\begin{equation}
  p(x) = a\,\frac{1 - x^{n+1}}{1-x}.
\end{equation}

\paragraph{Repeated squaring.}
Write $s_k$ for the upper-right entry of $L^{2^k}$. Since $L^1$ has upper-right entry $a$, we have $s_0 = a$. Squaring the matrix gives
\begin{equation}
  L^{2^{k+1}} = (L^{2^k})^2
  = \begin{pmatrix} x^{2^k} & s_k \\ 0 & 1 \end{pmatrix}^2
  = \begin{pmatrix} x^{2^{k+1}} & s_k(1 + x^{2^k}) \\ 0 & 1 \end{pmatrix},
\end{equation}
so the upper-right entry satisfies the recurrence
\begin{equation}
  s_{k+1} = s_k\bigl(1 + x^{2^k}\bigr).
\end{equation}
This is exactly Newton's method for computing $a/(1-x)$. Newton's iteration for the reciprocal $r = 1/(1-x)$ is $r_{k+1} = r_k\bigl(2 - (1-x)\,r_k\bigr)$. Starting from $r_0 = 1$, so that $s_k = a\,r_k$, we have $(1-x)\,r_k = 1 - x^{2^k}$ and therefore $2 - (1-x)\,r_k = 1 + x^{2^k}$, which reproduces the same recurrence. Each squaring step doubles the number of correct terms in the partial sum, just as each Newton step doubles the number of correct digits.

Although these two viewpoints give identical results in exact arithmetic, they differ in floating point. Both methods need the factor $1 + x^{2^k}$ at each step, and both compute it to the same absolute precision (about $\epsilon$ relative to the dominant term~$1$), so the \emph{per-step} rounding error is comparable. The difference is in how errors \emph{propagate} across steps.

For matrix squaring, if $\hat s_k = s_k(1 + \delta_k)$ carries relative error $\delta_k$, then the multiplication $\hat s_{k+1} = \hat s_k \cdot \mathrm{fl}(1 + \hat u_k)$ gives $\delta_{k+1} \approx \delta_k + O(\epsilon)$: the existing error propagates with factor~$1$. After $K$ steps, the accumulated relative error is $O(K\epsilon)$.

For Newton's iteration, linearizing the error $e_k = \hat r_k - r_k^*$ around the exact iterate $r_k^* = (1-x^{2^k})/(1-x)$ gives
\begin{equation}
  e_{k+1} = 2x^{2^k}\, e_k - (1-x)\,e_k^2 + O(\epsilon\, r_k^*).
\end{equation}
The key quantity is the derivative of the Newton map at $r_k^*$: since $f(r) = r(2 - (1-x)r)$, we have $f'(r_k^*) = 2 - 2(1-x)r_k^* = 2x^{2^k}$. In terms of relative error:
\begin{equation}
  \delta_{k+1} \approx \frac{2x^{2^k}}{1 + x^{2^k}}\,\delta_k + O(\epsilon).
\end{equation}

The stability comparison between the two methods depends on $|x|$. For $|x| < 1$, the propagation factor $2x^{2^k}/(1 + x^{2^k})$ has magnitude strictly less than~$1$ and shrinks to zero double-exponentially, so existing errors are \emph{damped} at each step. Errors from early steps decay geometrically, and the total accumulated error after $K$ steps is $O(\epsilon)$ independent of~$K$. Newton's iteration is more stable than matrix squaring in this regime.

For $|x| > 1$, the situation reverses. Now $x^{2^k}$ grows double-exponentially and $2x^{2^k}/(1 + x^{2^k}) \to 2$, so the propagation factor approaches~$2$ and existing errors roughly double at each step. After $K$ steps, errors from the earliest steps have been amplified by $\sim 2^K$. The finite partial sum $1 + x + \cdots + x^{2^K - 1}$ is a perfectly well-defined polynomial for any~$x$, but Newton's iteration computes it with exponentially growing error, while matrix squaring still accumulates error only as $O(K\epsilon)$. Matrix squaring is more stable in this regime.

The crossover is at $|x| = 1$, where both methods accumulate $O(K\epsilon)$ error. Overall, the nonlinear Newton form provides error damping inside the unit disk at the cost of error amplification outside it, while the linear matrix form has uniform $O(K\epsilon)$ accumulation everywhere. This contrasts with the continued fraction setting, where the Riccati projection can be \emph{less} stable (backward recurrence with parasitic solutions). The difference reflects the matrix structure: the upper-triangular polynomial matrices have no parasitic solution to excite.

\subsection*{Rounding error}

In exact arithmetic, the forward and backward recurrences give identical results. In floating point, they differ slightly. Horner's method (backward recurrence) uses $n$ multiplications and $n$ additions, while the forward recurrence uses $2n$ multiplications and $n$ additions---the extra $n$ multiplications maintain the running power $x^k$. This leads to error bounds of roughly
\begin{equation}
  |\hat p - p(x)| \lesssim n\epsilon\,\tilde p(|x|) \quad\text{(backward)}, \qquad
  |\hat p - p(x)| \lesssim 2n\epsilon\,\tilde p(|x|) \quad\text{(forward)},
\end{equation}
where $\tilde p(|x|) = \sum_k |a_k|\,|x|^k$ and $\epsilon$ is machine epsilon. Both methods are backward stable: the computed result is the exact evaluation of a polynomial with slightly perturbed coefficients. The difference is just a constant factor.

The relative error for both methods is controlled by the condition number $\tilde p(|x|)/|p(x)|$, which measures the cancellation between terms. When all terms have the same sign---for instance, when $x > 0$ and all $a_k > 0$---the condition number is~$1$ and both methods are equally accurate. Large relative errors require cancellation between terms, regardless of the method used.

The situation is very different for other bases. In a three-term recurrence like the Chebyshev recurrence $T_{n+1}(x) = 2x\,T_n(x) - T_{n-1}(x)$, the recurrence is second order and has two linearly independent solutions. For $|x| > 1$, one solution grows exponentially and the other decays. Forward recurrence is genuinely unstable in this setting: roundoff error excites the growing parasitic solution, producing errors that grow exponentially with~$n$. Backward recurrence (Clenshaw's algorithm, or more generally Olver's algorithm) avoids this by naturally selecting the minimal solution. The monomial-basis recurrence is effectively first order, so there is no parasitic solution to excite, and the stability difference between forward and backward evaluation remains modest.

\section{Continued fractions}

Now consider a continued fraction
\begin{equation}
  a_0 + \cfrac{b_1}{a_1 + \cfrac{b_2}{a_2 + \cfrac{b_3}{a_3 + \cdots}}}.
\end{equation}
Each level of the nesting corresponds to a fractional linear transformation $z \mapsto a_n + b_{n+1}/z$, represented by the matrix
\begin{equation}
  L_n = \begin{pmatrix} a_n & b_{n+1} \\ 1 & 0 \end{pmatrix}.
\end{equation}
The $n$th convergent of the continued fraction is then given by acting on $\binom{1}{0}$ on the right:
\begin{equation}
  (L_0 L_1 \cdots L_n) \begin{pmatrix} 1 \\ 0 \end{pmatrix} = \begin{pmatrix} p_n \\ q_n \end{pmatrix},
\end{equation}
which represents the value $p_n/q_n$. Writing
\begin{equation}
  L_0 L_1 \cdots L_n = \begin{pmatrix} p_n & p_n' \\ q_n & q_n' \end{pmatrix},
\end{equation}
the columns of the accumulated product have a direct interpretation: $p_n/q_n$ is the $n$th convergent, and $p_n'/q_n'$ is the value obtained by truncating the continued fraction one step earlier (i.e.\ by setting $b_{n+1} = 0$ instead of evaluating $\cdot\,\infty$).

For the common case of a simple continued fraction where all $b_n = 1$, each matrix takes the form
\begin{equation}
  L_n = \begin{pmatrix} a_n & 1 \\ 1 & 0 \end{pmatrix}.
\end{equation}

The parallel with polynomial evaluation is exact: both reduce to computing a product of $2 \times 2$ matrices and then extracting the result by acting on a distinguished point ($0$ for polynomials, $\infty$ for continued fractions). Associativity of matrix multiplication gives the same freedom to reorganize the computation in either case.

One difference is worth noting. In the polynomial case, the maps $z \mapsto a_k + xz$ are affine: every $L_k$ has bottom row $(0, 1)$, so the accumulated product is upper triangular and the denominator is always~$1$. This is what allows the sandwiched form $\begin{pmatrix} 1 & 0 \end{pmatrix} L_0 \cdots L_n \binom{0}{1} = p(x)$ to yield a single number. In the continued fraction case, the maps $z \mapsto a_n + b_{n+1}/z$ are genuinely projective: the denominator $q_n$ is nontrivial, so the result is inherently a ratio $p_n / q_n$ and no choice of row vector on the left can collapse it to a scalar.

\subsection*{Forward recurrence}

Accumulating the matrix product from left to right gives the \emph{forward recurrence} (or Wallis recurrence). Since the second column of the accumulated product satisfies $p_k' = b_{k+1}\, p_{k-1}$ (and similarly for $q$), appending $L_{k+1}$ yields
\begin{equation}
  p_{k+1} = a_{k+1}\, p_k + b_{k+1}\, p_{k-1}, \qquad
  q_{k+1} = a_{k+1}\, q_k + b_{k+1}\, q_{k-1},
\end{equation}
with initial values $p_{-1} = 1$, $p_0 = a_0$, $q_{-1} = 0$, $q_0 = 1$. This pair of linear recurrences computes the numerator and denominator of each successive convergent without any division; the ratio $p_n/q_n$ is formed only at the end.

\subsection*{Right-to-left accumulation and direct evaluation}

Now consider accumulating the matrix product from right to left. Define the \emph{tail product}
\begin{equation}
  R_k = L_k\, L_{k+1} \cdots L_n,
\end{equation}
built up by prepending one matrix at a time: $R_n = L_n$ and $R_k = L_k\, R_{k+1}$. Acting on $\infty$, the first column of $R_k$ gives the value of the tail continued fraction starting at index $k$. In particular, $R_0 \cdot \infty = p_n/q_n$ is the same $n$th convergent computed by the forward recurrence.

There is a different way to compute these tail values that does not use matrices at all. \emph{Direct evaluation} works from the inside out as a scalar recurrence:
\begin{equation}
  t_n = a_n, \qquad t_k = a_k + \frac{b_{k+1}}{t_{k+1}} \quad (k = n{-}1, \ldots, 0).
\end{equation}
Each step involves a division, and the final value $t_0$ is the $n$th convergent.

Right-to-left matrix accumulation and direct evaluation compute the same sequence of tail values, but they organize the arithmetic differently. Writing the first column of $R_k$ as $\binom{\alpha_k}{\beta_k}$, the recurrence $R_k = L_k\, R_{k+1}$ gives
\begin{equation}
  \alpha_k = a_k\,\alpha_{k+1} + b_{k+1}\,\beta_{k+1}, \qquad \beta_k = \alpha_{k+1}.
\end{equation}
Eliminating $\beta$, this becomes the three-term linear recurrence
\begin{equation}
  \alpha_k = a_k\,\alpha_{k+1} + b_{k+1}\,\alpha_{k+2},
\end{equation}
run backward from the initial values $\alpha_{n+1} = 1$, $\alpha_n = a_n$. This is the same form of recurrence as the Wallis recurrence for $p_k$ and $q_k$, but applied to different initial conditions and traversed in the opposite direction. The tail value at step $k$ is the ratio $t_k = \alpha_k / \alpha_{k+1}$.

This reveals the structural relationship between the two methods. The substitution $t_k = \alpha_k / \alpha_{k+1}$ converts the three-term linear recurrence for $\alpha$ into the nonlinear recurrence $t_k = a_k + b_{k+1}/t_{k+1}$: the per-step division in direct evaluation is exactly what converts a second-order linear recurrence into a first-order Riccati-type recurrence. Right-to-left matrix accumulation defers this division, carrying $\alpha_k$ and $\alpha_{k+1}$ as separate entries and forming their ratio only at the end.

\subsection*{Equivalence transformations}

An \emph{equivalence transformation} replaces the coefficients $a_k, b_k$ of a continued fraction with new coefficients $\tilde{a}_k, \tilde{b}_k$ while preserving every convergent. The classical result states that for any sequence of nonzero constants $c_1, c_2, \ldots$, the substitution
\begin{equation}
  \tilde{a}_k = c_k\, a_k \quad (k \geq 1), \qquad
  \tilde{b}_k = c_{k-1}\, c_k\, b_k \quad (k \geq 1),
\end{equation}
(with $c_0 = 1$ and $\tilde{a}_0 = a_0$) yields a continued fraction with the same convergents as the original.

In the matrix representation, this has a clean explanation: equivalence transformations correspond to inserting factorizations of the identity between terms in the matrix product. Let
\begin{equation}
  D_k = \begin{pmatrix} d_k & 0 \\ 0 & 1 \end{pmatrix},
\end{equation}
with $d_k \neq 0$, and insert $D_k D_k^{-1} = I$ between $L_{k-1}$ and $L_k$:
\begin{equation}
  L_0 L_1 \cdots L_n
  = (L_0\, D_1)(D_1^{-1} L_1\, D_2)(D_2^{-1} L_2\, D_3) \cdots (D_n^{-1} L_n).
\end{equation}
Computing the sandwiched factors gives
\begin{equation}
  D_k^{-1} L_k\, D_{k+1}
  = \begin{pmatrix} 1/d_k & 0 \\ 0 & 1 \end{pmatrix}
    \begin{pmatrix} a_k & b_{k+1} \\ 1 & 0 \end{pmatrix}
    \begin{pmatrix} d_{k+1} & 0 \\ 0 & 1 \end{pmatrix}
  = \begin{pmatrix} a_k\, d_{k+1}/d_k & b_{k+1}/d_k \\ d_{k+1} & 0 \end{pmatrix}.
\end{equation}
This is not quite in the standard continued fraction form $\bigl(\begin{smallmatrix} \tilde{a} & \tilde{b} \\ 1 & 0 \end{smallmatrix}\bigr)$, because the lower-left entry is $d_{k+1}$ rather than~$1$. But as a fractional linear transformation---acting on a homogeneous vector on the right---a matrix and any nonzero scalar multiple of it represent the same map. Dividing through by $d_{k+1}$ gives the projectively equivalent matrix
\begin{equation}
  \frac{1}{d_{k+1}}\begin{pmatrix} a_k\, d_{k+1}/d_k & b_{k+1}/d_k \\ d_{k+1} & 0 \end{pmatrix}
  = \begin{pmatrix} a_k/d_k & b_{k+1}/(d_k\, d_{k+1}) \\ 1 & 0 \end{pmatrix},
\end{equation}
which is in standard form with $\tilde{a}_k = a_k / d_k$ and $\tilde{b}_{k+1} = b_{k+1}/(d_k\, d_{k+1})$. Setting $c_k = 1/d_k$ recovers the classical equivalence transformation: $\tilde{a}_k = c_k\, a_k$ and $\tilde{b}_{k+1} = c_k\, c_{k+1}\, b_{k+1}$.

Why does this preserve all convergents? The $m$th convergent is obtained from the partial product $\tilde{L}_0 \cdots \tilde{L}_m$ acting on $\infty$. Before rescaling, the inserted identities telescope, so this partial product equals $L_0 L_1 \cdots L_m\, D_{m+1}$ times an overall scalar (which does not affect the fractional linear transformation). The partial product therefore differs from the original only by a trailing factor of $D_{m+1}$. But diagonal matrices of this form fix $\infty$:
\begin{equation}
  D_{m+1} \cdot \infty
  = \begin{pmatrix} d_{m+1} & 0 \\ 0 & 1 \end{pmatrix} \cdot \infty
  = \frac{d_{m+1} \cdot \infty}{0 \cdot \infty + 1}
  = \infty,
\end{equation}
so the trailing $D_{m+1}$ is invisible when we extract the convergent, and $(L_0 \cdots L_m\, D_{m+1}) \cdot \infty = (L_0 \cdots L_m) \cdot \infty$.

More generally, any upper triangular matrix fixes $\infty$, so one could insert upper triangular factorizations of the identity and still preserve all convergents. However, only diagonal insertions produce sandwiched factors that are projectively equivalent to continued fraction matrices (with lower-left entry~$1$ and lower-right entry~$0$). Upper triangular insertions would take us outside the world of continued fractions into more general compositions of fractional linear transformations.

\subsection*{The equivalent series}

The forward recurrence computes the convergent $f_n = p_n/q_n$ by tracking $p_n$ and $q_n$ through a pair of linear recurrences. But the convergent can also be expressed as a \emph{telescoping sum} of successive differences:
\begin{equation}
  f_n = a_0 + \sum_{k=1}^{n} \delta_k, \qquad \delta_k = f_k - f_{k-1}.
\end{equation}
Each term $\delta_k$ has a closed form that comes directly from the matrix representation.

\paragraph{The determinant identity.}
Since $\det(L_k) = a_k \cdot 0 - b_{k+1} \cdot 1 = -b_{k+1}$, the determinant of the accumulated product is
\begin{equation}
  \det(L_0 L_1 \cdots L_n) = \prod_{k=0}^{n} (-b_{k+1}) = (-1)^{n+1}\, b_1 b_2 \cdots b_{n+1}.
\end{equation}
Writing $M_n = \bigl(\begin{smallmatrix} p_n & p_n' \\ q_n & q_n' \end{smallmatrix}\bigr)$ and using $p_n' = b_{n+1}\, p_{n-1}$, $q_n' = b_{n+1}\, q_{n-1}$, we get
\begin{equation}
  \det(M_n) = p_n\, q_n' - p_n'\, q_n = b_{n+1}(p_n\, q_{n-1} - p_{n-1}\, q_n),
\end{equation}
which gives
\begin{equation}\label{eq:det-identity}
  p_n\, q_{n-1} - p_{n-1}\, q_n = (-1)^{n+1}\, b_1 b_2 \cdots b_n.
\end{equation}
This is the Euler--Minding identity. Dividing by $q_n\, q_{n-1}$, we obtain the convergent differences:
\begin{equation}
  \delta_n = f_n - f_{n-1} = \frac{(-1)^{n+1}\, b_1 b_2 \cdots b_n}{q_n\, q_{n-1}}.
\end{equation}
The continued fraction is therefore equal to the series
\begin{equation}
  f = a_0 + \sum_{n=1}^{\infty} \frac{(-1)^{n+1}\, b_1 b_2 \cdots b_n}{q_n\, q_{n-1}}.
\end{equation}

\paragraph{The Riccati projection, again.}
The ratio of successive differences satisfies
\begin{equation}
  \frac{\delta_n}{\delta_{n-1}}
  = \frac{-b_n}{q_n\, q_{n-1}} \cdot q_{n-1}\, q_{n-2}
  = -b_n\,\frac{q_{n-2}}{q_n}.
\end{equation}
Introducing the denominator ratio $D_n = q_{n-1}/q_n$, this becomes
\begin{equation}
  \delta_n = -b_n\, D_{n-1}\, D_n \cdot \delta_{n-1}.
\end{equation}
Since $q_n = a_n\, q_{n-1} + b_n\, q_{n-2}$, the ratio $D_n$ satisfies the Riccati recurrence
\begin{equation}
  D_n = \frac{1}{a_n + b_n\, D_{n-1}}, \qquad D_0 = 0.
\end{equation}
This is the same Riccati projection we saw earlier for the backward recurrence: the substitution $t_k = \alpha_k/\alpha_{k+1}$ converted the second-order linear recurrence into a first-order nonlinear one. Here, the substitution $D_n = q_{n-1}/q_n$ does the same for the forward Wallis recurrence.

Putting these together gives the \emph{forward series recurrence} (Steed's algorithm):
\begin{equation}
  \begin{aligned}
    D_n &= \frac{1}{a_n + b_n\, D_{n-1}}, \qquad D_0 = 0, \\[4pt]
    \delta_n &= -b_n\, D_{n-1}\, D_n \cdot \delta_{n-1}, \qquad \delta_1 = b_1\, D_1, \\[4pt]
    f_n &= f_{n-1} + \delta_n, \qquad f_0 = a_0.
  \end{aligned}
\end{equation}
Each step requires only scalar arithmetic (no matrix storage), avoids the overflow and underflow issues of the Wallis recurrence (since $D_n$ and $\delta_n$ remain moderate in size), and has a built-in convergence test: stop when $|\delta_n|$ is small enough. A closely related variant is the modified Lentz algorithm, which tracks the numerator ratio $C_n = p_n/p_{n-1} = a_n + b_n/C_{n-1}$ alongside $D_n$ and uses the multiplicative update $f_n = f_{n-1} \cdot C_n D_n$.

\paragraph{Relationship to equivalence transformations.}
The $D_n$ recurrence is itself a M\"obius iteration---the map $z \mapsto 1/(a_n + b_n z)$ is represented by the matrix $G_n = \bigl(\begin{smallmatrix} 0 & 1 \\ b_n & a_n \end{smallmatrix}\bigr)$---so the denominator ratio $D_n$ is computed by its own product of $2 \times 2$ matrices, just a different set of matrices from the $L_k$. The equivalent series thus rests on two purely matrix-theoretic ingredients: the \emph{determinant identity} (the determinant of a product is the product of the determinants) and the \emph{Riccati projection} (converting the second-order linear recurrence into a first-order nonlinear one by taking ratios of successive terms).

Neither ingredient is an equivalence transformation in the sense of the preceding section. Equivalence transformations modify the $L_k$ factors while preserving convergents; they use predetermined constants $c_k$ that are independent of the particular continued fraction being evaluated. The forward series recurrence instead extracts the convergent from an \emph{unmodified} matrix product by tracking the denominator ratios $D_n$, which depend on the accumulated product itself. It is closer in spirit to an LU decomposition---a data-dependent factorization---than to a change of variables with fixed parameters.

\section{The Riccati projection}

A common pattern runs through both the polynomial and continued fraction settings: a matrix representation carries more state than the scalar quantity we ultimately want, and eliminating the extra state introduces nonlinearity. We have now seen several instances of this, and it is worth pausing to describe the pattern explicitly.

\subsection*{Continued fractions}

The Wallis recurrence tracks the pair $(p_n, q_n)$ through a linear system:
\begin{equation}
  p_{n+1} = a_{n+1}\, p_n + b_{n+1}\, p_{n-1}, \qquad
  q_{n+1} = a_{n+1}\, q_n + b_{n+1}\, q_{n-1}.
\end{equation}
Both components satisfy the same three-term recurrence; they differ only in initial conditions. The convergent $f_n = p_n/q_n$ is a single number extracted from the pair by taking a ratio---the natural projection from homogeneous coordinates to the projective line. Substituting $f_n = p_n/q_n$ into the recurrence converts the linear system in two variables into the nonlinear Riccati recurrence in one:
\begin{equation}
  t_k = a_k + \frac{b_{k+1}}{t_{k+1}},
\end{equation}
for the backward direction (direct evaluation), or the analogous recurrence $D_n = 1/(a_n + b_n\, D_{n-1})$ for the forward direction (the denominator ratio in Steed's algorithm). The per-step division is the price of collapsing two linear state variables into one nonlinear one.

\subsection*{Polynomials and Newton's method}

The same pattern appears in the polynomial setting, though with a different source of redundancy. Repeated squaring of $L = \bigl(\begin{smallmatrix} x & a \\ 0 & 1 \end{smallmatrix}\bigr)$ tracks the pair $(u_k, s_k)$ where $u_k = x^{2^k}$ and $s_k$ is the partial sum $a(1 + x + \cdots + x^{2^k - 1})$. The squaring recurrence is
\begin{equation}
  u_{k+1} = u_k^2, \qquad s_{k+1} = s_k\,(1 + u_k).
\end{equation}
The first equation is quadratic but self-contained: $u_k$ evolves independently. Given $u_k$, the second equation is \emph{linear} in $s_k$. The upper-triangular structure of $L$ makes the system triangularly coupled rather than fully coupled.

The partial sum formula $s_k = a(1 - u_k)/(1-x)$ gives an algebraic relation between the two state variables: $u_k = 1 - (1-x)\, r_k$ where $r_k = s_k/a$. Substituting into $s_{k+1} = s_k(1 + u_k)$ eliminates $u_k$:
\begin{equation}
  r_{k+1} = r_k\bigl(2 - (1-x)\, r_k\bigr),
\end{equation}
which is Newton's iteration for the reciprocal of $1-x$. The quadratic nonlinearity arises because the bilinear coupling $s_k \cdot u_k$ becomes $r_k \cdot g(r_k)$ when $u_k$ is expressed in terms of $r_k$.

\subsection*{The common structure}

In both cases, the matrix representation carries an auxiliary quantity that keeps the recurrence for the quantity of interest simple:
\begin{center}
\begin{tabular}{lll}
  & \textbf{Two-variable (matrix)} & \textbf{One-variable (scalar)} \\[3pt]
  CF & Wallis: linear in $(p_n, q_n)$ & Riccati: rational in $f_n$ \\
  Polynomial & squaring: linear in $s_k$ given $u_k$ & Newton: quadratic in $r_k$
\end{tabular}
\end{center}
The auxiliary variable is $q_n$ in the continued fraction case and $u_k = x^{2^k}$ in the polynomial case. Eliminating it---by forming a ratio $p/q$ or by substituting $u = 1 - (1-x)r$---introduces nonlinearity.

The difference between the two cases reflects the matrix structure. The continued fraction matrices $\bigl(\begin{smallmatrix} a_n & b_{n+1} \\ 1 & 0 \end{smallmatrix}\bigr)$ live in $\mathrm{GL}(2)$: both rows of the accumulated product carry nontrivial information, and the projection is the ratio of homogeneous coordinates. The polynomial matrices $\bigl(\begin{smallmatrix} x & a \\ 0 & 1 \end{smallmatrix}\bigr)$ live in the upper-triangular (affine) subgroup: the bottom row is fixed at $(0,1)$, so there is no projective ratio to form. Instead, the redundancy is algebraic---the two nontrivial entries of the accumulated product are related by the partial sum formula---and the projection is an algebraic substitution rather than a homogeneous ratio.

\end{document}
